---
title: "Blending visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Blending visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
# Sharp figures: rendered by ragg at 120 dpi, so a full-width figure is still
# drawn with more pixels than the page shows. JPEG at quality 80 keeps detailed
# rasters and photos small without visible artefacts; the palette chart, flat
# colours and fine text, overrides it with PNG. knitr knows ragg_png by name but
# not a ragg JPEG device, hence this wrapper.
ragg_jpeg <- function(filename, width, height, ...) {
  ragg::agg_jpeg(filename, width = width, height = height, units = "in",
                 res = knitr::opts_current$get("dpi"), quality = 80)
}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  fig.width = 7, fig.height = 6, dpi = 120,
  dev = "ragg_jpeg", fig.ext = "jpeg",
  panels = TRUE
)
# Narrow margins and plain, understated panel titles for every figure, set out
# of sight so the visible code shows only what matters for blending.
knitr::knit_hooks$set(panels = function(before, options, envir) {
  if (before) graphics::par(mar = c(0.3, 0.3, 1.6, 0.3), font.main = 1,
                            cex.main = 1.05, col.main = "grey20")
})
```

```{r setup, include = FALSE}
library(rvtr)
# The sample data is downloaded on demand. If it cannot be fetched while the
# vignette builds, every chunk below is shown without being run.
pt <- c(9.9464, 51.6317)
has_data <- tryCatch({ rvt_data_lgln(pt, "dtm"); TRUE }, error = function(e) FALSE)
knitr::opts_chunk$set(eval = has_data)
```

```{r offline-note, eval = !has_data, echo = FALSE, results = "asis"}
cat("> The sample data could not be downloaded when this vignette was built,",
    "so the code is shown without its figures. Run it yourself to see them.\n")
```

No single visualization shows everything about a landscape. A hillshade gives
the eye a familiar sense of relief but hides whatever runs parallel to its
light. A sky-view factor shows enclosed ground but flattens slopes. An
orthophoto shows what is there, but not the shape of the ground beneath it.
The usual answer is to lay several of them over one another, and
`rvt_blend()` does that with the same blend modes a GIS offers.

This vignette explains how a blend is built, shows all thirteen modes side by
side, works through two composites and ends with the ready-made composite
the package provides.

## The sample data

`rvt_data_lgln()` reads open data for any place in Lower Saxony, streamed from
the state survey authority. A point returns the 1 km tile it lies in; here,
the tile beside Burg Hardenberg: farmland, woodland, a castle on its spur and a
village. The orthophoto is fetched at 1 m, so it lies on the terrain grid.

```{r data, fig.height = 2.55}
pt  <- c(9.9464, 51.6317)                              # longitude, latitude
dtm <- rvt_data_lgln(pt, "dtm")                        # terrain: the bare ground
rgb <- rvt_data_lgln(pt, "rgb", year = 2025, res = 1)  # orthophoto at 1 m

hs  <- rvt_hillshade(dtm)
svf <- rvt_svf(dtm, reach = 10)

op <- par(mfrow = c(1, 3))
rvt_plot(rgb, main = "Orthophoto")
rvt_plot(hs,  main = "Hillshade")
rvt_plot(svf, main = "Sky-view factor")
par(op)
```

Each layer is downloaded the first time it is requested and read from a local
cache afterwards.

Data: © GeoBasis-DE/LGLN `r format(Sys.Date(), "%Y")`, CC BY 4.0, Daten
geändert.

## How a blend is built

### The raster you pipe in is the background

`rvt_blend()` lays the raster passed as its argument *over* the raster piped
into it. A chain therefore reads from the bottom up, the way a GIS layer panel
does:

```{r order}
stack <- hs |>
  rvt_blend(svf, "multiply", opacity = 0.5, range = c(0.7, 1))
```

Order matters, because several modes treat their two inputs differently.
`subtract` removes the layer from the background, so swapping the two gives a
very different picture:

```{r asymmetry, fig.height = 3.75}
op <- par(mfrow = c(1, 2))
hs |>
  rvt_blend(svf, "subtract", opacity = 0.5, range = c(0.7, 1)) |>
  rvt_plot(main = "Hillshade minus SVF")
rvt_stack(svf, range = c(0.7, 1)) |>
  rvt_blend(hs, "subtract", opacity = 0.5) |>
  rvt_plot(main = "SVF minus hillshade")
par(op)
```

Both are drawn at half opacity. At full strength the left image would be
almost entirely black, because the stretched sky-view factor (SVF) is brighter
than the hillshade over nearly the whole tile, and a subtraction cannot go
below black.

`overlay`, `hard_light`, `dodge` and `burn` depend on order as well, although
often less visibly. `overlay`, for instance, only changes when the layers are
swapped where one of them is darker than mid-grey and the other lighter.

### Nothing is computed until you render

`rvt_blend()` returns a recipe, not a raster. Printing it lists the layers
from the bottom up:

```{r print}
stack
```

However long the chain, the layers are read and blended in a single pass when
the result is needed. `rvt_plot()` draws a stack directly, and `rvt_render()`
writes it to a file:

```{r render, eval = FALSE}
rvt_render(stack, "relief.tif")
```

Computing the metrics is the expensive part of any composite; blending them is
cheap. Because each metric is computed once and kept, trying another mode or
opacity takes seconds.

### Every layer is stretched first

Before blending, each layer is stretched to the range 0 to 1. By default the
stretch runs from the raster's own minimum to its maximum, which often wastes
contrast. A sky-view factor can in principle take any value from 0 to 1, but
on ordinary terrain almost all of it lies near the top. `range` focuses the
stretch on the values that actually occur, and `rvt_range()` reads a
percentile range from the raster:

```{r ranges}
rvt_range(svf)
rvt_range(svf, pct = c(2, 98))
```

The range is fixed before any part of the raster is read for blending, so the
result does not depend on how the raster is divided into tiles. `invert = TRUE`
flips a layer after stretching, the usual way to show slope.

### Opacity is applied last

The mode combines the layer with what lies beneath it, and `opacity` then
fades that result back towards the background. At `opacity = 0` the layer has
no effect; at `1` the mode applies at full strength.

## All thirteen modes

Every mode below is applied to the same two layers: the hillshade as the
background, and the sky-view factor over it at full opacity.

```{r modes, fig.height = 5}
op <- par(mfrow = c(3, 5))
rvt_plot(hs,  main = "Background", max_dim = 400)
rvt_plot(svf, main = "Layer", max_dim = 400)
for (mode in rvt_blend_modes) {
  hs |>
    rvt_blend(svf, mode, range = c(0.7, 1)) |>
    rvt_plot(main = mode, max_dim = 400)
}
par(op)
```

| Mode | What it does | Use it for |
|---|---|---|
| `normal` | The layer replaces the background | Fading one view into another with `opacity` |
| `multiply` | Darkens by the layer, never lightens | Deepening enclosed ground with a sky-view factor |
| `screen` | Lightens by the layer, never darkens | Lifting ridges, or brightening a dark base |
| `overlay` | Darkens the dark and lightens the light parts of the background | Adding contrast without washing the image out |
| `soft_light` | A gentler `overlay` | Subtle contrast, or relief over an orthophoto |
| `hard_light` | `overlay`, but driven by the layer | Letting the layer's contrast dominate |
| `darken` | Keeps the darker of the two | Combining the shadows of two views |
| `lighten` | Keeps the lighter of the two | Combining the highlights of two views |
| `difference` | The absolute difference | Showing where two views disagree |
| `subtract` | Background minus layer, limited at black | Removing one view's brightness from another |
| `addition` | The sum, limited at white | Brightening quickly; clips easily |
| `dodge` | Brightens the background by the layer | Strong highlights; clips easily |
| `burn` | Darkens the background by the layer | Strong shadows; clips easily |

In practice, three of these cover most composites: `multiply`, `screen` and
`overlay`.

## Recipes

### A relief stack from terrain metrics

This stack combines four views of the terrain. The hillshade gives the basic
impression of relief, inverted slope sharpens breaks of slope, positive
openness blended with `overlay` lifts convex features, and the sky-view factor
multiplied in darkens enclosed ground.

```{r relief, fig.height = 3.75}
slp  <- rvt_slope(dtm)
opns <- rvt_openness(dtm, reach = 10)

relief <- function(slope_range, openness_range, svf_range) {
  rvt_stack(hs, range = c(0, 1)) |>
    rvt_blend(slp,  "normal",   opacity = 0.5,  range = slope_range, invert = TRUE) |>
    rvt_blend(opns, "overlay",  opacity = 0.5,  range = openness_range) |>
    rvt_blend(svf,  "multiply", opacity = 0.25, range = svf_range)
}

op <- par(mfrow = c(1, 2))
relief(c(0, 50), c(68, 93), c(0.7, 1)) |> rvt_plot(main = "General ranges")
relief(c(0, 15), c(85, 93), c(0.9, 1)) |> rvt_plot(main = "Narrow ranges")
par(op)
```

The left image uses the layers, modes and ranges of the general-terrain half
of `rvt_vat()`. Those ranges suit rugged ground, and on this gently rolling
tile they leave the picture pale. The right image narrows them to the
flat-terrain values, which spends the contrast on the gentle slopes that
actually occur here. Written out as a stack, every layer, mode and range can be
changed individually.

### An orthophoto with shaded relief

Laying terrain over imagery shows colour and shape together. `multiply` only
darkens, so slopes facing away from the light become darker while sunlit ones
keep the photo's own brightness. `soft_light` works in both directions,
darkening shaded slopes and lightening sunlit ones.

```{r ortho, fig.height = 3.75}
op <- par(mfrow = c(1, 2))
rgb |> rvt_blend(hs, "multiply", opacity = 0.6) |> rvt_plot(main = "Multiply at 60%")
rgb |> rvt_blend(hs, "soft_light") |> rvt_plot(main = "Soft light")
par(op)
```

A single-band layer is applied to each of the orthophoto's three colour bands.
The two can be blended directly because, fetched at 1 m, the orthophoto lies on
exactly the terrain grid.

## Ready-made composites

`rvt_vat()` is a composite that needs no assembling: the Visualization for
Archaeological Topography of Kokalj and Somrak. It builds the four-layer relief
stack from the recipe above twice, once with settings for ordinary relief and
once tuned for flat ground, and averages the two, so a single call works across
mixed terrain. Passing the same preset twice renders one terrain type alone.

```{r vat, fig.height = 2.55}
general <- list(rvt_preset_general, rvt_preset_general)
flat    <- list(rvt_preset_flat, rvt_preset_flat)

op <- par(mfrow = c(1, 3))
rvt_vat(dtm) |> rvt_plot(main = "VAT combined")
rvt_vat(dtm, presets = general) |> rvt_plot(main = "VAT general")
rvt_vat(dtm, presets = flat) |> rvt_plot(main = "VAT flat")
par(op)
```

The two presets differ in more than their stretch ranges. The flat-terrain
version also lowers the sun from 35 to 15 degrees, searches twice as far for
openness and sky-view factor, and removes more noise. Both presets are
ordinary lists, so a variant is one `modifyList()` away:

```{r vat-custom, eval = FALSE}
wider <- modifyList(rvt_preset_general, list(reach = 20))
rvt_vat(dtm, presets = list(wider, rvt_preset_flat))
```

A ready-made composite is the quick route. When a layer needs to change in a
way the presets do not expose, write the stack out with `rvt_blend()`, as in
the relief recipe above.

## Choosing colours

`rvt_plot()` takes a palette name as its second argument, as in
`rvt_plot(svf, "Viridis")`. `rvt_palettes()` draws every available palette, so
you can choose one by eye:

```{r palettes, fig.height = 6.5, dev = "ragg_png", fig.ext = "png"}
rvt_palettes("sequential")
```
