---
title: "Blending visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Blending visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
source("_common.R")
```

```{r setup, include = FALSE}
library(rvtr)
# The sample data is downloaded on demand. If it cannot be fetched while the
# vignette builds, every chunk below is shown without being run.
pt <- c(9.9464, 51.6317)
has_data <- tryCatch({ rvt_data_lgln(pt, "dtm"); TRUE }, error = function(e) FALSE)
knitr::opts_chunk$set(eval = has_data)
```

```{r offline-note, eval = !has_data, echo = FALSE, results = "asis"}
cat("> The sample data could not be downloaded when this vignette was built,",
    "so the code is shown without its figures. Run it yourself to see them.\n")
```

No single visualization shows everything about a landscape. A hillshade gives
the eye a familiar sense of relief but hides whatever runs parallel to its
light. A sky-view factor shows enclosed ground but flattens slopes. An
orthophoto shows what is there, but not the shape of the ground beneath it.
The usual answer is to lay several of them over one another, and
`rvt_blend()` does that with the same blend modes a GIS offers.

This vignette explains how a blend is built, shows all thirteen modes side by
side, works through three composites, covers the ready-made composite the
package provides, and ends by relighting an orthophoto with its terrain.

## The sample data

`rvt_data_lgln()` reads open data for any place in Lower Saxony, streamed from
the state survey authority. A point returns the 1 km tile it lies in; here,
the tile beside Burg Hardenberg: farmland, woodland, a castle on its spur and a
village. Three products are used: the orthophoto, fetched at 1 m so it lies on
the terrain grid; the terrain model, the bare ground; and the surface model,
the ground with trees and buildings on it. The [data vignette](data.html)
covers fetching data in detail.

Below, both elevation models are shown with the anisotropic sky-view factor
(`rvt_asvf()`), which gives a sense of light from the north-west while hiding
nothing in shadow. The terrain model reveals the castle's spur, the slopes and
straight lines running beneath the forest; the surface model shows the tree
crowns and houses that cover them.

```{r data, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
pt  <- c(9.9464, 51.6317)                              # longitude, latitude
rgb <- rvt_data_lgln(pt, "rgb", year = 2025, res = 1)  # orthophoto at 1 m
dtm <- rvt_data_lgln(pt, "dtm")                        # terrain: the bare ground
dsm <- rvt_data_lgln(pt, "dsm")                        # surface: trees and buildings

rvt_plot(rgb, main = "Orthophoto", max_dim = 400)
rvt_asvf(dtm, reach = 10) |> rvt_plot(main = "Terrain model", max_dim = 400,
                                      minmax_def = c(0.7, 1))
rvt_asvf(dsm, reach = 10) |> rvt_plot(main = "Surface model", max_dim = 400,
                                      minmax_def = c(0.4, 1))
```

Data: © GeoBasis-DE/LGLN `r format(Sys.Date(), "%Y")`, CC BY 4.0, Daten
geändert.

Each layer is downloaded the first time it is requested and read from a local
cache afterwards. The blends below start from two views of the terrain model,
computed once:

```{r metrics}
hs  <- rvt_hillshade(dtm)
svf <- rvt_svf(dtm, reach = 10)
```

## How a blend is built

### The raster you pipe in is the background

`rvt_blend()` lays the raster passed as its argument *over* the raster piped
into it. A chain therefore reads from the bottom up, the way a GIS layer panel
does:

```{r order}
stack <- hs |>
  rvt_blend(svf, "multiply", opacity = 0.5, range = c(0.7, 1))
```

Order matters, because several modes treat their two inputs differently.
`subtract` removes the layer from the background, so swapping the two gives a
very different picture:

```{r asymmetry, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
hs |>
  rvt_blend(svf, "subtract", opacity = 0.5, range = c(0.7, 1)) |>
  rvt_plot(main = "Hillshade minus SVF", max_dim = 400)
rvt_stack(svf, range = c(0.7, 1)) |>
  rvt_blend(hs, "subtract", opacity = 0.5) |>
  rvt_plot(main = "SVF minus hillshade", max_dim = 400)
```

Both are drawn at half opacity. At full strength the left image would be
almost entirely black, because the stretched sky-view factor (SVF) is brighter
than the hillshade over nearly the whole tile, and a subtraction cannot go
below black.

`overlay`, `hard_light`, `dodge` and `burn` depend on order as well, although
often less visibly. `overlay`, for instance, only changes when the layers are
swapped where one of them is darker than mid-grey and the other lighter.

### Nothing is computed until you render

`rvt_blend()` returns a recipe, not a raster. Printing it lists the layers
from the bottom up:

```{r print}
stack
```

However long the chain, the layers are read and blended in a single pass when
the result is needed. `rvt_plot()` draws a stack directly, and `rvt_render()`
writes it to a file:

```{r render, eval = FALSE}
rvt_render(stack, "relief.tif")
```

Computing the metrics is the expensive part of any composite; blending them is
cheap. Because each metric is computed once and kept, trying another mode or
opacity takes seconds.

### Every layer is stretched first

Before blending, each layer is stretched to the range 0 to 1. By default the
stretch runs from the raster's own minimum to its maximum, which often wastes
contrast. A sky-view factor can in principle take any value from 0 to 1, but
on ordinary terrain almost all of it lies near the top. `range` focuses the
stretch on the values that actually occur, and `rvt_range()` reads a
percentile range from the raster:

```{r ranges}
rvt_range(svf)
rvt_range(svf, pct = c(2, 98))
```

The range is fixed before any part of the raster is read for blending, so the
result does not depend on how the raster is divided into tiles. `invert = TRUE`
flips a layer after stretching, the usual way to show slope.

### Opacity is applied last

The mode combines the layer with what lies beneath it, and `opacity` then
fades that result back towards the background. At `opacity = 0` the layer has
no effect; at `1` the mode applies at full strength.

## All thirteen modes

Every mode below is applied to the same two layers: the hillshade as the
background, and the sky-view factor over it at full opacity.

```{r modes, fig.width = fig_w(5, 240), fig.height = fig_h(3, 240), layout = c(3, 5)}
rvt_plot(hs,  main = "Background", max_dim = 240)
rvt_plot(svf, main = "Layer", max_dim = 240)
for (mode in rvt_blend_modes) {
  hs |>
    rvt_blend(svf, mode, range = c(0.7, 1)) |>
    rvt_plot(main = mode, max_dim = 240)
}
```

| Mode | What it does | Use it for |
|---|---|---|
| `normal` | The layer replaces the background | Fading one view into another with `opacity` |
| `multiply` | Darkens by the layer, never lightens | Deepening enclosed ground with a sky-view factor |
| `screen` | Lightens by the layer, never darkens | Lifting ridges, or brightening a dark base |
| `overlay` | Darkens the dark and lightens the light parts of the background | Adding contrast without washing the image out |
| `soft_light` | A gentler `overlay` | Subtle contrast, or relief over an orthophoto |
| `hard_light` | `overlay`, but driven by the layer | Letting the layer's contrast dominate |
| `darken` | Keeps the darker of the two | Combining the shadows of two views |
| `lighten` | Keeps the lighter of the two | Combining the highlights of two views |
| `difference` | The absolute difference | Showing where two views disagree |
| `subtract` | Background minus layer, limited at black | Removing one view's brightness from another |
| `addition` | The sum, limited at white | Brightening quickly; clips easily |
| `dodge` | Brightens the background by the layer | Strong highlights; clips easily |
| `burn` | Darkens the background by the layer | Strong shadows; clips easily |

In practice, three of these cover most composites: `multiply`, `screen` and
`overlay`.

## Recipes

### A relief stack from terrain metrics

This stack combines four views of the terrain. The hillshade gives the basic
impression of relief, inverted slope sharpens breaks of slope, positive
openness blended with `overlay` lifts convex features, and the sky-view factor
multiplied in darkens enclosed ground.

```{r relief, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
slp  <- rvt_slope(dtm)
opns <- rvt_openness(dtm, reach = 10)

relief <- function(slope_range, openness_range, svf_range) {
  rvt_stack(hs, range = c(0, 1)) |>
    rvt_blend(slp,  "normal",   opacity = 0.5,  range = slope_range, invert = TRUE) |>
    rvt_blend(opns, "overlay",  opacity = 0.5,  range = openness_range) |>
    rvt_blend(svf,  "multiply", opacity = 0.25, range = svf_range)
}

relief(c(0, 50), c(68, 93), c(0.7, 1)) |> rvt_plot(main = "General ranges", max_dim = 400)
relief(c(0, 15), c(85, 93), c(0.9, 1)) |> rvt_plot(main = "Narrow ranges", max_dim = 400)
```

The left image uses the layers, modes and ranges of the general-terrain half
of `rvt_vat()`. Those ranges suit rugged ground, and on this gently rolling
tile they leave the picture pale. The right image narrows them to the
flat-terrain values, which spends the contrast on the gentle slopes that
actually occur here. Written out as a stack, every layer, mode and range can be
changed individually.

### An orthophoto with shaded relief

Laying terrain over imagery shows colour and shape together. `multiply` only
darkens, so slopes facing away from the light become darker while sunlit ones
keep the photo's own brightness. `soft_light` works in both directions,
darkening shaded slopes and lightening sunlit ones.

```{r ortho, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rgb |> rvt_blend(hs, "multiply", opacity = 0.6) |> rvt_plot(main = "Multiply at 60%", max_dim = 400)
rgb |> rvt_blend(hs, "soft_light") |> rvt_plot(main = "Soft light", max_dim = 400)
```

A single-band layer is applied to each of the orthophoto's three colour bands.
The two can be blended directly because, fetched at 1 m, the orthophoto lies on
exactly the terrain grid.

### Terrain and surface together

The terrain model shows the ground and the surface model what stands on it,
and two blends show both at once.

The first finds the height of everything above the ground. Stretching both
elevation models over the same range, here 300 m, and subtracting the terrain
from the surface leaves exactly that height, as a share of the range: 0.1 is
30 m. Rendered to a raster, it shows the forest, the houses and the single
trees along the field, while open ground stays black.

The second uses that raster as a layer. Inverted and multiplied over the
anisotropic sky-view factor of the terrain, it lays a grey tint wherever
something stands, darker the taller it is. The shape of the ground stays sharp
underneath, so the castle's spur, the slopes and the lines beneath the forest
remain as readable as in the terrain model alone.

```{r terrain-surface, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
same <- c(120, 420)                             # one range for both models
height <- rvt_stack(dsm, range = same) |>
  rvt_blend(dtm, "subtract", range = same) |>
  rvt_render()
rvt_plot(height, main = "Height above ground", max_dim = 400, minmax_def = c(0, 0.1))

rvt_stack(rvt_asvf(dtm, reach = 10), range = c(0.7, 1)) |>
  rvt_blend(height, "multiply", opacity = 0.6, range = c(0, 0.1), invert = TRUE) |>
  rvt_plot(main = "Ground tinted by what stands on it", max_dim = 400)
```

## Ready-made composites

`rvt_vat()` is a composite that needs no assembling: the Visualization for
Archaeological Topography of Kokalj and Somrak. It builds the four-layer relief
stack from the recipe above twice, once with settings for ordinary relief and
once tuned for flat ground, and averages the two, so a single call works across
mixed terrain. Passing the same preset twice renders one terrain type alone.

```{r vat, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
general <- list(rvt_preset_general, rvt_preset_general)
flat    <- list(rvt_preset_flat, rvt_preset_flat)

rvt_vat(dtm) |> rvt_plot(main = "VAT combined", max_dim = 400)
rvt_vat(dtm, presets = general) |> rvt_plot(main = "VAT general", max_dim = 400)
rvt_vat(dtm, presets = flat) |> rvt_plot(main = "VAT flat", max_dim = 400)
```

The two presets differ in more than their stretch ranges. The flat-terrain
version also lowers the sun from 35 to 15 degrees, searches twice as far for
openness and sky-view factor, and removes more noise. Both presets are
ordinary lists, so a variant is one `modifyList()` away:

```{r vat-custom, eval = FALSE}
wider <- modifyList(rvt_preset_general, list(reach = 20))
rvt_vat(dtm, presets = list(wider, rvt_preset_flat))
```

A ready-made composite is the quick route. When a layer needs to change in a
way the presets do not expose, write the stack out with `rvt_blend()`, as in
the relief recipe above.

## Relighting an orthophoto

Blending lays shading over a photo. `rvt_relight()` goes a step further and
lights the photo with the terrain beneath it, the way a rendered landscape
model is lit: a warm sun on slopes that face it, soft cast shadows, and cool
sky light that gives the shadows a slight blue cast. Using the surface model
lets trees and buildings cast shadows too.

The result is darker than the photo wherever something casts a shadow. On this
tile, where trees and buildings cover more than half the ground, the whole
image comes out at about two thirds of the photo's brightness, while sunlit
open ground keeps nearly all of it.

```{r relight, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
true_colour <- c(0, 0, 0, 255, 255, 255)

rvt_plot(rgb, main = "Orthophoto", minmax_def = true_colour, max_dim = 400)
rvt_relight(rgb, dsm) |> rvt_plot(main = "Relit", minmax_def = true_colour, max_dim = 400)
```

The lighting comes in named styles, listed in `rvt_relight_styles`. Any
setting passed directly overrides the style, so
`rvt_relight(rgb, dsm, style = "winter", sun_elevation = 10)` gives a winter
light with an even lower sun.

```{r relight-styles, fig.width = fig_w(3), fig.height = fig_h(2), layout = c(2, 3)}
styles <- c("verdant", "golden_hour", "winter",
            "twilight", "blue_hour", "vivid")

for (style in styles) {
  rvt_relight(rgb, dsm, style = style) |>
    rvt_plot(main = style, minmax_def = true_colour, max_dim = 400)
}
```

## Choosing colours

`rvt_plot()` takes a palette name as its second argument, as in
`rvt_plot(svf, "Viridis")`. `rvt_palettes()` draws every available palette, so
you can choose one by eye:

```{r palettes, fig.width = px(1212), fig.height = px(1125), dev = "ragg_png", fig.ext = "png"}
rvt_palettes("sequential")
```
