---
title: "Getting data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
source("_common.R")
```

```{r setup, include = FALSE}
library(rvtr)
# Everything here except the last section is fetched on demand. If either
# source cannot be reached while the vignette builds, those chunks are shown
# without being run.
has_data <- tryCatch({
  rvt_data_lgln(c(9.9464, 51.6317), "dtm")
  rvt_data_mapterhorn(c(8.2618, 48.7788), res = 30, size = 1000)
  TRUE
}, error = function(e) FALSE)
knitr::opts_chunk$set(eval = has_data)
```

```{r offline-note, eval = !has_data, echo = FALSE, results = "asis"}
cat("> The data could not be downloaded when this vignette was built, so the",
    "code is shown without its results. Run it yourself to see them.\n")
```

Every function in this package takes the path to an elevation raster, so any
terrain model you already have works. Two functions fetch one for you:
`rvt_data_mapterhorn()` for anywhere in the world, and `rvt_data_lgln()` for
Lower Saxony, where it also offers a surface model and orthophotos. Neither
downloads anything until the data is read, and then only the part that is
needed.

Distances in this package are in metres, never pixels. Both functions
therefore return data in a coordinate system measured in metres, and your own
data needs one too.

## Terrain for anywhere: `rvt_data_mapterhorn()`

[Mapterhorn](https://mapterhorn.com) combines open elevation data into one
global dataset. The 30 m Copernicus model covers the whole world, and national
LiDAR terrain models take over where they have been published: 1 m across
Germany, Austria and Norway, 0.5 m in Switzerland, 0.25 m in parts of
Baden-Württemberg, and many more.

Give a point, and you get a square around it, `size` metres across. Give an
extent, and you get exactly that area. Here is one place, the Battert crags
above Baden-Baden, at three scales, each shown with `rvt_vat()`. Each panel is
400 × 400 pixels, and each outline marks the area of the panel before it.

```{r scales, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
battert <- c(8.2618, 48.7788)                            # longitude, latitude

fine   <- rvt_data_mapterhorn(battert, size = 100)            # finest: 0.25 m
medium <- rvt_data_mapterhorn(battert, size = 800,   res = 2)
coarse <- rvt_data_mapterhorn(battert, size = 12000, res = 30)

# VAT searches 10 and 20 m around each cell, less than one 30 m cell,
# so the coarse panel searches further
wide <- list(modifyList(rvt_preset_general, list(reach = 300)),
             modifyList(rvt_preset_flat,    list(reach = 600)))

outline <- function(inner) {                             # the area of `inner`
  b <- methods::new(gdalraster::GDALRaster, inner)$bbox()
  rect(b[1], b[2], b[3], b[4], border = "#d7301f", lwd = 2)
}

rvt_vat(fine) |> rvt_plot(main = "0.25 m over 100 m", max_dim = 400)
rvt_vat(medium) |> rvt_plot(main = "2 m over 800 m", max_dim = 400)
outline(fine)
rvt_vat(coarse, presets = wide) |> rvt_plot(main = "30 m over 12 km", max_dim = 400)
outline(medium)
```

At 0.25 m single sandstone blocks stand out on the forested slope. At 2 m the
line of crags appears along the ridge, with forest roads and skid trails
winding around it, and at 30 m the ridge is one of many among the valleys at
the western edge of the Black Forest.

### Which data covers a place

`rvt_data_mapterhorn_sources()` lists the sources available for an area,
finest first, without fetching any elevation data:

```{r sources}
rvt_data_mapterhorn_sources(battert)[, c("source", "resolution", "license")]
```

### Resolution

Without `res`, you get the finest source available. A larger `res` coarsens
the data, and a smaller one is refused, because the result could only be
interpolated:

```{r too-fine, error = TRUE}
rvt_data_mapterhorn(battert, size = 100, res = 0.1)
```

Where an area crosses the edge of a detailed survey, the coarser data fills in
wherever the finer data stops, so the result has no holes.

### Coordinate system

Mapterhorn stores its tiles in Web Mercator, whose map units are stretched
by 1/cos(latitude), 1.6 times at 51 degrees north. That would break every
distance in this package, so the data is always reprojected: by default to the
UTM zone of the area's centre, or to any coordinate system in metres you pass
as `crs`. Use the national system when the data has to line up with other
rasters, for example `crs = 25832` in Germany.

```{r crs}
gdalraster::srs_find_epsg(methods::new(gdalraster::GDALRaster, fine)$getProjection())
```

### Credit

The data comes from many producers, each under its own open licence. The
sources used are printed the first time each is fetched in a session, and
<https://mapterhorn.com/attribution> lists the full terms.

## Lower Saxony in detail: `rvt_data_lgln()`

The state survey authority of Lower Saxony (LGLN) publishes three products for
the whole state, and `rvt_data_lgln()` reads all of them:

| product | what | native grid |
|---|---|---|
| `"dtm"` | terrain model, the bare ground | 1 m |
| `"dsm"` | surface model, ground plus trees and buildings | 1 m |
| `"rgb"` | orthophoto | 0.2 m |

The data comes in the survey's own coordinate system, ETRS89 / UTM zone 32N
(EPSG:25832), and is never reprojected. A point returns the 1 km tile it lies
in; an extent, here a 400 m square in EPSG:25832, returns exactly that area.
Fetching the orthophoto at `res = 1` puts it on the same grid as the terrain,
so the three line up cell for cell.

```{r lgln, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
area <- sf::st_bbox(c(xmin = 565300, ymin = 5720300, xmax = 565700, ymax = 5720700),
                    crs = 25832)                         # beside Burg Hardenberg

dtm <- rvt_data_lgln(area, "dtm")
dsm <- rvt_data_lgln(area, "dsm")
rgb <- rvt_data_lgln(area, "rgb", year = 2025, res = 1)

rvt_vat(dtm) |> rvt_plot(main = "Terrain model", max_dim = 400)
rvt_vat(dsm) |> rvt_plot(main = "Surface model", max_dim = 400)
rvt_plot(rgb, main = "Orthophoto", max_dim = 400,
         minmax_def = c(0, 0, 0, 255, 255, 255))
```

The terrain model shows the shape of the ground, even beneath the forest; the
surface model adds the tree crowns and buildings standing on it.

### Flight years

The terrain and surface models usually exist for one year, but orthophotos are
flown every few years. `rvt_data_lgln_years()` lists them without fetching any
data:

```{r years}
rvt_data_lgln_years(area, "rgb")
```

Without `year`, the most recent year covering the whole area is used. The
comparison below is an 80 m square at the orthophoto's native 0.2 m, so each
panel shows its own 400 × 400 pixels. Flights happen in different seasons:
2013 was flown in June, with the trees in leaf, and 2025 in March, without.

```{r years-plot, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
detail <- sf::st_bbox(c(xmin = 565460, ymin = 5720460, xmax = 565540, ymax = 5720540),
                      crs = 25832)

for (year in c(2013, 2025)) {
  rvt_data_lgln(detail, "rgb", year = year) |>
    rvt_plot(main = year, max_dim = 400, minmax_def = c(0, 0, 0, 255, 255, 255))
}
```

### Credit

LGLN's data is licensed CC BY 4.0. Credit it as © GeoBasis-DE/LGLN followed by
the year you downloaded it, and add "Daten geändert" (data modified) for
anything derived from it, which includes every result of this package.

## Streaming and downloading

Both functions return the path to a small virtual raster by default. It points
at the remote files, and data moves only when something reads it, so asking
for an area costs nothing until you compute on it.

With `download = TRUE` a Cloud-Optimized GeoTIFF is stored in your user cache
instead, and later calls for the same area are served from there. Use
`refresh = TRUE` to fetch it again.

```{r download, eval = FALSE}
dtm <- rvt_data_mapterhorn(battert, size = 2000, download = TRUE)
tools::R_user_dir("rvtr", "cache")                       # where it is kept
```

Download first when you will compute several metrics over a large area. Each
metric reads the data in many small windows, and over the network that is far
slower than reading a local file.

Before anything is read, both functions estimate the download and stop above
`max_mb`, 500 MB by default:

```{r guard, error = TRUE}
rvt_data_mapterhorn(c(8.5, 48.2, 9.5, 48.6))             # 75 x 45 km at 0.25 m
```

## Your own data

Any raster GDAL can read works, as long as its coordinate system is measured
in metres. Reproject one in degrees first:

```{r reproject, eval = FALSE}
gdalraster::warp("dem_wgs84.tif", "dem_utm.tif", t_srs = "EPSG:25832")
```

Terrain often comes as many tiles. Pass them all at once; they are read as one
seamless raster, so results show no edges at the tile boundaries:

```{r mosaic, eval = FALSE}
tiles <- list.files("dtm_tiles", "\\.tif$", full.names = TRUE)
tiles |> rvt_svf("svf.tif")
```

`rvt_fill()` interpolates across holes in the data. Fill before computing
anything: slope and curvature lose a ring of cells around every hole, and the
horizon-based metrics see straight through one.

`rvt_resample()` coarsens a raster, to study terrain at a broader scale or to
make a large area affordable. It refuses to make a raster finer, since that
could only interpolate. Here part of the bundled sample tile is shown at its
own 1 m and coarsened to 4 m, where the small bumps and faint paths disappear:

```{r own, eval = TRUE, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
dem <- system.file("extdata", "dtm1.tif", package = "rvtr")
part <- tempfile(fileext = ".tif")
gdalraster::translate(dem, part, cl_arg = c("-srcwin", 300, 300, 400, 400), quiet = TRUE)

rvt_vat(part) |> rvt_plot(main = "1 m", max_dim = 400, minmax_def = c(0, 1))
rvt_resample(part, 4) |> rvt_vat() |>
  rvt_plot(main = "Coarsened to 4 m", max_dim = 400, minmax_def = c(0, 1))
```
