---
title: "Visualizations and metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizations and metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
source("_common.R")
```

```{r setup, include = FALSE}
library(rvtr)
# The terrain is downloaded on demand. If it cannot be fetched while the
# vignette builds, every chunk below is shown without being run.
area <- sf::st_bbox(c(xmin = 572492, ymin = 5714542, xmax = 572892, ymax = 5714942),
                    crs = 25832)
has_data <- tryCatch({ rvt_data_lgln(area, "dtm", download = TRUE); TRUE },
                     error = function(e) FALSE)
knitr::opts_chunk$set(eval = has_data)
```

```{r offline-note, eval = !has_data, echo = FALSE, results = "asis"}
cat("> The terrain could not be downloaded when this vignette was built, so the",
    "code is shown without its figures. Run it yourself to see them.\n")
```

This vignette shows every visualization and metric in the package on the same
ground, so they can be compared directly. Each function takes the path to a
terrain model and writes a raster, and every distance it takes is in metres.
The short descriptions here are a starting point; each help page explains how
the method works, how to tune it, and which settings suit which terrain, and
`?rvtr` summarises which to use where.

The terrain is a 400 m square of the Hünstollen at 1 m, from the terrain model
of Lower Saxony. A steep, curving edge drops from a field on higher ground in
the south-west, crossed by ditches, to lower and uneven ground in the east, and
tracks run across the slopes. The area is wooded, but a terrain model shows the
bare ground beneath the trees. For the landscape around it, a 4 km square
centred on the same place is fetched at 10 m, which also gives 400 × 400 cells.
Both are downloaded once, since every figure below reads them:

```{r data}
area <- sf::st_bbox(c(xmin = 572492, ymin = 5714542, xmax = 572892, ymax = 5714942),
                    crs = 25832)
dem <- rvt_data_lgln(area, "dtm", download = TRUE)

landscape_area <- sf::st_bbox(c(xmin = 570690, ymin = 5712740,
                                xmax = 574690, ymax = 5716740), crs = 25832)
landscape <- rvt_data_lgln(landscape_area, "dtm", res = 10, download = TRUE)

outline <- function() {                     # the 400 m square, on a landscape map
  rect(area[["xmin"]], area[["ymin"]], area[["xmax"]], area[["ymax"]], lwd = 2)
}
```

Each figure's code shows the palette and stretch used, as a starting point for
your own. The panels are 400 × 400 pixels: one pixel per metre for the 400 m
square, one per 10 m for the landscape.

## Shading and sun

`rvt_hillshade()` lights the terrain from one direction and is the most
familiar picture of relief. Anything running parallel to the light fades out,
and slopes facing away from it turn dark. `rvt_multi_hillshade()` lights from
four directions at once, so less is hidden and no side is left in deep shadow.
`rvt_shadow()` shows true cast shadow for one sun position: white where the sun
reaches, black where terrain blocks it.

All three light the terrain from the north-west by default, here with the sun
25 degrees high for the shadow. The sun never stands there in Central Europe;
the north-west is a display convention, because relief lit from the top of the
image reads correctly, while light from below makes hills look like hollows.

```{r shading, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
rvt_hillshade(dem) |> rvt_plot(main = "rvt_hillshade()", max_dim = 400)
rvt_multi_hillshade(dem) |> rvt_plot(main = "rvt_multi_hillshade()", max_dim = 400)
rvt_shadow(dem, sun_elevation = 25) |> rvt_plot(main = "rvt_shadow()", max_dim = 400)
```

`rvt_daylight()` and `rvt_insolation()` use the real path of the sun through
the year instead. `rvt_daylight()` counts the hours of direct sun per day, here
averaged over a year, taking both cast shadow and the way each slope faces into
account. `rvt_insolation()` measures the solar energy received, in kWh/m² per
year. Hours and energy are not the same thing: a slope facing away from the
sun can be lit for many hours at a low angle while receiving little energy.

Both are theoretical. They assume a clear sky, and on a terrain model they see
only the bare ground: this area is wooded, so in reality far less sun reaches
the ground than shown. For the light falling on the treetops, use a surface
model instead.

```{r sun, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_daylight(dem) |>
  rvt_plot("Inferno", main = "rvt_daylight()", max_dim = 400, range = c(5, 12))
rvt_insolation(dem) |>
  rvt_plot("Inferno", main = "rvt_insolation()", max_dim = 400, range = c(600, 2000))
```

## Sky and horizon

These metrics look along many directions from each cell and find how high the
horizon rises. Nothing is lit from one side, so no feature disappears in
shadow.

`rvt_svf()`, the sky-view factor, is the share of the sky visible from each
cell: 1 on open flat ground, lower in hollows and at the foot of slopes. It
works like an overcast sky. `rvt_asvf()` weights the sky towards one direction,
giving a sense of light like a hillshade while still hiding nothing.
`rvt_sky_illumination()` models the diffuse light actually received, counting
the direction each slope faces as well as the horizon.

```{r sky, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
rvt_svf(dem) |> rvt_plot(main = "rvt_svf()", max_dim = 400, range = c(0.65, 1))
rvt_asvf(dem) |> rvt_plot(main = "rvt_asvf()", max_dim = 400, range = c(0.65, 1))
rvt_sky_illumination(dem) |>
  rvt_plot(main = "rvt_sky_illumination()", max_dim = 400, range = c(0.6, 1))
```

`rvt_openness()` measures the angle of the open sky above each cell, averaged
over all directions, in degrees. Convex features such as banks and edges come
out bright. `rvt_openness_negative()` does the same for the terrain turned
upside down, so ditches, tracks and hollows come out bright instead.

```{r openness, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_openness(dem) |>
  rvt_plot(main = "rvt_openness()", max_dim = 400, range = c(75, 92))
rvt_openness_negative(dem) |>
  rvt_plot(main = "rvt_openness_negative()", max_dim = 400, range = c(75, 92))
```

## Local relief

These metrics remove the broad shape of the landscape and keep what stands
above or sinks below it, in metres or in units of local variation. Signed
results read best with a diverging palette and a range centred on zero; below,
red is above the surroundings and blue below them.

`rvt_slrm()`, the simple local relief model, is each cell's height above or
below the average of its surroundings, here within 20 m; `rvt_tpi()` is the
same function under its other common name, the topographic position index.
`rvt_msrm()` combines this over a range of scales, so features of different
sizes show at once. `rvt_dev()` divides the difference by how rough the
surroundings are, so a small bank on smooth ground and a larger one on rough
ground read alike.

```{r relief, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
rvt_slrm(dem) |>
  rvt_plot("Blue-Red 3", main = "rvt_slrm()", max_dim = 400, range = c(-2, 2))
rvt_msrm(dem) |>
  rvt_plot("Blue-Red 3", main = "rvt_msrm()", max_dim = 400, range = c(-0.5, 0.5))
rvt_dev(dem) |>
  rvt_plot("Blue-Red 3", main = "rvt_dev()", max_dim = 400, range = c(-1, 1))
```

`rvt_local_dominance()` shows how far the surroundings fall away below a cell,
which brings out low, spread-out features. `rvt_mstp()` colours each cell by
the scale at which it stands out: blue for fine, green for medium, red for
broad features. Its default scales reach several kilometres, far beyond this
square, so the scales here are cut down to fit.

```{r relief2, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_local_dominance(dem) |>
  rvt_plot(main = "rvt_local_dominance()", max_dim = 400, range = c(0.5, 2.5))
rvt_mstp(dem, local = c(3, 21, 2), meso = c(23, 103, 18), broad = c(123, 223, 50)) |>
  rvt_plot(main = "rvt_mstp()", max_dim = 400)
```

## Surface shape

`rvt_slope()` is steepness in degrees, drawn here with flat ground light and
steep ground dark. `rvt_aspect()` is the compass direction each slope faces,
from 0 to 360 degrees, so it needs a palette whose ends meet. `rvt_curvature()`
is how sharply the ground bends: positive (red) where it is convex, such as the
top of an edge, negative (blue) where it is concave, such as the foot of a
slope. Measured from each cell's immediate neighbours at 1 m, it also picks up
every small bump, hence the speckled look; the sections below show what its
different types show, and how to measure it at a broader scale. `rvt_log()`,
the Laplacian of Gaussian, is curvature with smoothing built in: it first
smooths the ground by `sigma` metres, so edges of that size stand out and
smaller noise does not. Like curvature, it is positive on banks and edges and
negative in ditches and hollows.

```{r shape, fig.width = fig_w(2), fig.height = fig_h(2), layout = c(2, 2)}
rvt_slope(dem) |>
  rvt_plot(rev(hcl.colors(256, "Grays")), main = "rvt_slope()", max_dim = 400,
           range = c(0, 45))
rvt_aspect(dem) |>
  rvt_plot(rainbow(256), main = "rvt_aspect()", max_dim = 400, range = c(0, 360))
rvt_curvature(dem) |>
  rvt_plot("Blue-Red 3", main = "rvt_curvature()", max_dim = 400,
           range = c(-0.2, 0.2))
rvt_log(dem, sigma = 3) |>
  rvt_plot("Blue-Red 3", main = "rvt_log()", max_dim = 400, range = c(-0.1, 0.1))
```

### Landforms

`rvt_geomorphons()` sorts each cell into one of ten landform classes by
comparing it with the terrain along eight directions, up to `reach` metres
away. The classes are listed in `rvt_geomorphon_classes`; the colours below
follow the usual convention, from greys for flat ground and reds for peaks and
ridges through yellows for slopes to greens and blues for hollows, valleys and
pits.

Landforms are a matter of scale. On the 400 m square at 1 m, looking 20 m
around, almost everything is slope, with the ditches and parts of the edge
picked out as narrow lines. On the 4 km landscape at 10 m, looking 500 m around, the larger
forms appear: level plateaus in grey, their rims as shoulders, the ridge on
which the square lies, and the network of valleys cut into them. Here ground
counts as flat below 3 degrees rather than the default 1, because the plateau
tops are gently tilted.

```{r geomorphons, fig.width = fig_w(3), fig.height = fig_h(1), layout = c(1, 3)}
classes <- c("#dcdcdc", "#380000", "#c80000", "#ff5014", "#fad23c",
             "#ffff3c", "#b4e614", "#3cfa96", "#0000ff", "#000038")

rvt_geomorphons(dem, reach = 20) |>
  rvt_plot(classes, main = "1 m, reach = 20", max_dim = 400, range = c(1, 10))
rvt_geomorphons(landscape, reach = 500, skip = 20, flat_threshold = 3) |>
  rvt_plot(classes, main = "10 m, reach = 500", max_dim = 400, range = c(1, 10))
outline()

plot.new()
legend("center", legend = names(rvt_geomorphon_classes), fill = classes,
       bty = "n", cex = 1.1)
```

### Types of curvature

"Curvature" covers several different questions about how the ground bends, and
`rvt_curvature()` answers 14 of them through `type`. The six below are the ones
most often useful. They are measured over a 3 m window (`radius = 3`) to calm
the noise, and each has its own stretch, because their values differ in size.

- `profile`, the default, is measured straight down the slope: positive where
  the slope steepens towards the brow of an edge, negative where it eases off
  at its foot. It picks out breaks of slope, such as the banks of the ditches.
- `tangential` is measured across the slope, along the contour: positive on
  spurs, where water would spread apart, negative in hollows and gullies,
  where it gathers.
- `mean` averages the bending over every direction, so it shows whether the
  ground bulges outwards or dishes inwards overall.
- `minimal` and `maximal` are the least and the most the ground bends in any
  direction. Along a ditch the ground bends hard across it and barely along it,
  so ditches stand out in `minimal` (blue) and banks and edges in `maximal`
  (red).
- `shape_index` describes the form alone, however gently or sharply the ground
  bends: −1 for a pit, −0.5 a valley, 0 a saddle, +0.5 a ridge and +1 a dome.
  Because intensity is ignored, even the faintest bumps on flat ground show,
  which gives the busy texture.

```{r curvature-types, fig.width = fig_w(3), fig.height = fig_h(2), layout = c(2, 3)}
types <- c(profile = 0.1, tangential = 0.05, mean = 0.05,
           minimal = 0.1, maximal = 0.1, shape_index = 1)
for (type in names(types)) {
  rvt_curvature(dem, type = type, radius = 3) |>
    rvt_plot("Blue-Red 3", main = type, max_dim = 400,
             range = c(-types[[type]], types[[type]]))
}
```

The remaining types, among them `gaussian` (domes and bowls against saddles),
`unsphericity` and `twisting`, are described in `?rvt_curvature`.

## Choosing a scale

Every metric has a scale: the size of the features it responds to, set by the
cell size of the terrain model together with the window or distance the metric
looks at. The same metric at a different scale answers a different question.

### Local and landscape

Below, curvature is measured over the same window of 7 × 7 cells twice. On the
400 m square at 1 m (`radius = 3`), the window spans 7 m, and curvature finds
the banks of the ditches, the tracks and the sharp rim of the edge. On the 4 km
landscape at 10 m (`radius = 30`), the window spans 70 m, and those details are
gone: curvature now traces the valleys cut into the plateaus in blue, with
their rims in red. The outline marks the 400 m square. Values at the landscape scale
are ten times smaller, since curvature is measured per metre, so the stretch
differs too.

```{r scale, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_curvature(dem, radius = 3) |>
  rvt_plot("Blue-Red 3", main = "1 m cells, radius = 3", max_dim = 400,
           range = c(-0.1, 0.1))
rvt_curvature(landscape, radius = 30) |>
  rvt_plot("Blue-Red 3", main = "10 m cells, radius = 30", max_dim = 400,
           range = c(-0.008, 0.008))
outline()
```

A coarser terrain model is also far quicker to process, so for a landscape
question `rvt_resample()`, or fetching at a coarser `res`, is the efficient
route, rather than a very large window on fine data.

### Window size

Within one terrain model, `radius` sets the size of the surroundings a cell is
compared with. The local relief models show this well. A small `radius` picks out tracks, ditches and single mounds; a large one
shows whole slopes and edges:

```{r radius, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_slrm(dem, radius = 5) |>
  rvt_plot("Blue-Red 3", main = "rvt_slrm(radius = 5)", max_dim = 400,
           range = c(-0.5, 0.5))
rvt_slrm(dem, radius = 30) |>
  rvt_plot("Blue-Red 3", main = "rvt_slrm(radius = 30)", max_dim = 400,
           range = c(-4, 4))
```

### How far to look

The sky and horizon metrics, and the sun metrics, have `reach` instead: how far
they look for the horizon, in metres. A short reach shows small features close
by; a long reach adds the shape of the wider landscape: the lower ground below
the edge and in the east darkens, because the higher ground around it closes
off the horizon. A long reach stays affordable, because distant terrain
is read from coarser copies of the terrain model; `?rvt_reach` explains how.

```{r reach, fig.width = fig_w(2), fig.height = fig_h(1), layout = c(1, 2)}
rvt_openness(dem, reach = 10) |>
  rvt_plot(main = "rvt_openness(reach = 10)", max_dim = 400, range = c(75, 92))
rvt_openness(dem, reach = 100) |>
  rvt_plot(main = "rvt_openness(reach = 100)", max_dim = 400, range = c(70, 90))
```

## Composites

No single metric shows everything, and they combine well. `rvt_vat()` builds a
ready-made composite from a hillshade, slope, openness and sky-view factor, and
`rvt_blend()` lets you build your own; the
[blending vignette](blending.html) covers both.
